Algorithms for the Automated Correction of Vertical Drift in Eye Tracking Data

### 2020-07-25 manuscript/manuscript.tex
Added discussion of Yamaya et al

A number of methods have previously been developed to automate post-hoc vertical drift correction. \texttt{FixAlign}, an R package developed by \textcite{Cohen:2013}, is currently the most well-established method in the experimental psychology community, although other methods have recently been proposed by \textcite{Schroeder:0000} and \textcite{Spakov:2019}. In addition, there is a disparate body of work from several subfields of computer science, such as biometrics \parencite{Abdulin:2015}, educational technology \parencite{Hyrskykari:2006}, and user-interface design \parencite{Beymer:2005}, in which various ad-hoc algorithms have been reported \parencite[see also][]{LimaSanches:2015,Carl:2013,Mishra:2012,MartinezGomez:2012}.

@@

The \texttt{compare} algorithm is based on the method reported by \textcite{LimaSanches:2015}. The fixation sequence is first segmented into ``gaze lines'' by identifying the return sweeps---long saccades that move the eye from the end of one line to the start of the next. The algorithm considers any saccade that moves from right to left by more than some threshold value (we use 512~px) to be a return sweep. Gaze lines are then matched to text lines based on a measure of similarity between them.

\textcite{LimaSanches:2015} considered three measures of similarity and found dynamic time warping \parencite[DTW;][]{Sakoe:1978} to be the best method. DTW returns a measure of how dissimilar two sequences are, so the \texttt{compare} algorithm simply matches each gaze line to whichever text line results in the smallest DTW distance. In the example in Fig.~\ref{fig02}D, we only depict the first gaze line (black dots), which is compared to each line of text; the best match is with the first text line, as reflected in its low cost (i.e., the sum of the mapping lines is minimized).

The gaze lines and text lines are only compared in terms of their \textit{x}~values under the assumption that the fixations in a gaze line should have a good horizontal alignment with the centers of the words in the corresponding text line. Relying only on the \textit{x}~values helps the algorithm overcome vertical drift issues, but it is also problematic because the lines of text in a passage tend to be horizontally similar to each other; each line tends to contain a similar number of words that are of a similar length, resulting in potential ambiguity about how gaze lines and text lines should be matched up. To alleviate this issue, \textcite{LimaSanches:2015} only compare the gaze line to a certain number of nearby text lines, controlled by a free parameter (we use 3).

@@

The final algorithm we consider, \texttt{warp}, is also novel to this paper, although it is mostly a wrapper around a preexisting algorithm---dynamic time warping. DTW was used by \textcite{LimaSanches:2015} to provide a measure of dissimilarity between a gaze line and a text line in the \texttt{compare} algorithm. To our knowledge, however, there have been no previous reports of DTW being used for its primary purpose: finding the best alignment between two time sequences. This is somewhat surprising because DTW is the natural computational choice for tackling vertical drift. Two methods have previously been described that share some conceptual similarities. \textcite{Carl:2013} uses a basket of reading-related measures to place a cost on different paths through a lattice of fixation-to-character mappings and selects the path with minimal cost. \textcite{MartinezGomez:2012} convert the fixation sequence into a low resolution image and find the best transformation of this image to fit the overall shape of the passage. Both of these methods are quite convoluted, however, and we consider them to be special cases of \texttt{warp}, which is a direct application of the standard DTW algorithm to eye tracking data.

@@

As with the simulations, accuracy is measured as the percentage of fixations that the algorithm mapped to the correct line; the ground truth is defined by the gold standard manual correction. In cases where the correctors chose to discard a fixation, the algorithm is automatically wrong. Fig.~\ref{fig08} plots accuracy on the 48 sample trials by algorithm. The most striking result is \texttt{compare} with overall median accuracy of 61.2\%, substantially worse than all other algorithms.\footnote{This concurs with the 60\% accuracy reported by \textcite[p.~1231]{LimaSanches:2015}.} This contrasts with our simulations, which indicated that \texttt{compare} should at~least be relatively strong on distortion. The reason for this discrepancy is that the simulated fixation sequences were generated directly from the lines of text with one fixation per word, so the artificial gaze lines that \texttt{compare} identified tended to be very horizontally similar to the artificial text lines. In the natural dataset, however, this is not the case; when the data contains a~lot of natural noise and regressions, gaze lines cannot be reliably matched to text lines based on similarity, even if the set of candidate text lines is narrowed down to the three closest neighbors. Given that \texttt{compare} exhibited such poor performance, we consider it to be an algorithmic deadend and do not discuss it any further.

>>>>

A number of methods have previously been developed to automate post-hoc vertical drift correction. \texttt{FixAlign}, an R package developed by \textcite{Cohen:2013}, is currently the most well-established method in the experimental psychology community, although other methods have recently been proposed by \textcite{Schroeder:0000} and \textcite{Spakov:2019}. In addition, there is a disparate body of work from several subfields of computer science, such as biometrics \parencite{Abdulin:2015}, educational technology \parencite{Hyrskykari:2006}, and user-interface design \parencite{Beymer:2005}, in which various ad-hoc algorithms have been reported \parencite[see also][]{LimaSanches:2015,Carl:2013,Mishra:2012,MartinezGomez:2012,Yamaya:2017}.

@@

The \texttt{compare} algorithm is directly based on the method reported by \textcite{LimaSanches:2015} and is very similar to the more complex methods described by \textcite{Yamaya:2017}. The fixation sequence is first segmented into ``gaze lines'' by identifying the return sweeps---long saccades that move the eye from the end of one line to the start of the next. The algorithm considers any saccade that moves from right to left by more than some threshold value (we use 512~px) to be a return sweep. Gaze lines are then matched to text lines based on a measure of similarity between them. \textcite{LimaSanches:2015} considered three measures of similarity and found dynamic time warping \parencite{Vintsyuk:1968,Sakoe:1978} to be the best method. Similarly, \textcite{Yamaya:2017} use the closely related Needleman--Wunsch algorithm \parencite{Needleman:1970}. In the example in Fig.~\ref{fig02}D, we only depict the first gaze line (black dots), which is compared to each line of text; the best match is with the first text line, as reflected in its low cost (i.e., the sum of the mapping lines is minimized).

The gaze lines and text lines are only compared in terms of their \textit{x}~values under the assumption that the fixations in a gaze line should have a good horizontal alignment with the centers of the words in the corresponding text line. Relying only on the \textit{x}~values helps the algorithm overcome vertical drift issues, but it is also problematic because the lines of text in a passage tend to be horizontally similar to each other; each line tends to contain a similar number of words that are of a similar length, resulting in potential ambiguity about how gaze lines and text lines should be matched up. To alleviate this issue, both \textcite{LimaSanches:2015} and \textcite{Yamaya:2017} only compare the gaze line to a certain number of nearby text lines (we set this parameter to 3, which is effectively the closest line plus one line above and one line below).

@@

The final algorithm we consider, \texttt{warp}, is also novel to this paper, although it is mostly a wrapper around a preexisting algorithm---dynamic time warping \parencite[DTW;][]{Vintsyuk:1968,Sakoe:1978}. DTW was used by the \texttt{compare} algorithm to provide a measure of dissimilarity between a gaze line and a text line. To our knowledge, however, there have been no previous reports of DTW being used for its primary purpose: finding the best alignment between two time sequences. This is somewhat surprising because DTW is the natural computational choice for tackling vertical drift. Two methods have previously been described that share some conceptual similarities. \textcite{Carl:2013} uses a basket of reading-related measures to place a cost on different paths through a lattice of fixation-to-character mappings and selects the path with minimal cost. \textcite{MartinezGomez:2012} convert the fixation sequence into a low resolution image and find the best transformation of this image to fit the overall shape of the passage. Both of these methods are quite convoluted, however, and we consider them to be special cases of \texttt{warp}, which is a direct application of the standard DTW algorithm to eye tracking data.

@@

As with the simulations, accuracy is measured as the percentage of fixations that the algorithm mapped to the correct line; the ground truth is defined by the gold standard manual correction. In cases where the correctors chose to discard a fixation, the algorithm is automatically wrong. Fig.~\ref{fig08} plots accuracy on the 48 sample trials by algorithm. The most striking result is \texttt{compare} with overall median accuracy of 61.2\%, substantially worse than all other algorithms.\footnote{This concurs with the 60\% accuracy reported by \textcite[p.~1231]{LimaSanches:2015}. \textcite[p.~104]{Yamaya:2017}, who describe a fairly similar, albeit more complex, method, report accuracy of 87\%, which is still quite low compared to the algorithms considered in this paper.} This contrasts with our simulations, which indicated that \texttt{compare} should at~least be relatively strong on distortion. The reason for this discrepancy is that the simulated fixation sequences were generated directly from the lines of text with one fixation per word, so the artificial gaze lines that \texttt{compare} identified tended to be very horizontally similar to the artificial text lines. In the natural dataset, however, this is not the case; when the data contains a~lot of natural noise and regressions, gaze lines cannot be reliably matched to text lines based on similarity, even if the set of candidate text lines is narrowed down to the three closest neighbors. Given that \texttt{compare} exhibited such poor performance, we consider it to be an algorithmic deadend and do not discuss it any further.
